---
title: "How does CmdStanR work?"
author: "Jonah Gabry and Rok Češnovar"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{How does CmdStanR work?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="children/_settings-knitr.Rmd"}
```

## Introduction

This vignette is intended to be read after the [_Getting started with CmdStanR_](https://mc-stan.org/cmdstanr/articles/cmdstanr.html)
vignette. Please read that first for important background. In this document we
provide additional details about compiling models, passing in data, and how
CmdStan output is saved and read back into R.

We will only use the `$sample()` method in examples, but all model fitting
methods work in a similar way under the hood.

```{r setup, message=FALSE}
library(cmdstanr)
check_cmdstan_toolchain(quiet = TRUE)
```

## Compilation

### Compiling a Stan program

The `cmdstan_model()` function creates a new `CmdStanModel` object, which stores
the path to a Stan program as well as the path to a compiled executable.

```{r compile}
# Copy the example to the temporary directory so the vignette doesn't modify
# the CmdStan installation
example_stan_file <- file.path(
  cmdstan_path(), "examples", "bernoulli", "bernoulli.stan"
)
stan_file <- file.path(tempdir(), "bernoulli.stan")
invisible(file.copy(example_stan_file, stan_file))

# Compile the model
mod <- cmdstan_model(stan_file)
mod$print()
mod$stan_file()
mod$exe_file()
```

Subsequently, if you create a `CmdStanModel` object from the same Stan file
then compilation will be skipped (assuming the file hasn't changed).

```{r already-compiled}
mod <- cmdstan_model(stan_file)
```

`cmdstan_model()` builds the executable before it returns, or reuses one that
is already up to date. The arguments that configure the build are passed to
`cmdstan_model()` directly, including `include_paths`, `cpp_options`, and 
various other arguments. See `?cmdstan_model` for details.

```{r compile-options, eval=FALSE}
mod <- cmdstan_model(
  stan_file,
  include_paths = "paths/to/directories/with/included/files",
  user_header = "path/to/header/file.hpp",
  cpp_options = list(stan_threads = TRUE)
)
```



### Compiling without a model object

`compile_stan_file()` runs the same build as `cmdstan_model()` but returns the
path to the executable instead of a `CmdStanModel` object. The path can be given
to `cmdstan_model(exe_file = )` later, which uses the executable as it is. One
use case for this is an R package that compiles its Stan programs when it is
installed and loads them when it runs.

```{r compile_stan_file}
unlink(mod$exe_file())
exe_file <- compile_stan_file(stan_file)
exe_file
mod_from_exe <- cmdstan_model(exe_file = exe_file)
```

The `unlink()` call above deleted the executable that `mod` was created with,
and `compile_stan_file()` built a new one in its place. A `CmdStanModel` object
only runs the executable it was created with, so `mod` would now refuse to run
and, since the rest of this vignette uses `mod`, we recreate it here. Since the
executable is up to date, it does not compile again. 

```{r recreate}
mod <- cmdstan_model(stan_file)
```

### Pedantic check

CmdStanR will always check that your Stan program does not contain any invalid
syntax but with `pedantic` mode enabled the check will also warn you about other
potential issues in your model, for example:

- Distribution usage issues: distribution arguments do not match the
distribution specification, or some specific distribution is used in an
inadvisable way.
- Unused parameter: a parameter is defined but does not contribute to target.
- Large or small constant in a distribution: very large or very small constants
are used as distribution arguments.
- Control flow depends on a parameter: branching control flow (like if/else)
depends on a parameter value.
- Parameter has multiple tildes: a parameter is on the left-hand side of
multiple tildes (i.e., multiple `~` symbols).
- Parameter has zero or multiple priors: a parameter has zero or more than one
prior distribution.
- Variable is used before assignment: a variable is used before being assigned a
value.
- Strict or nonsensical parameter bounds: a parameter is given questionable
bounds.
- Nonlinear transformations: a nonlinear transformation on the left-hand side
of a sampling statement may require a Jacobian adjustment.

For the latest information on the checks performed in pedantic mode see the
[Pedantic mode section](https://mc-stan.org/docs/stan-users-guide/using-stanc.html#pedantic-mode)
in the Stan User's Guide.

Pedantic mode is available when compiling the model or when using the separate
`$check_syntax()` method of a `CmdStanModel` object. Internally this corresponds
to setting the `stanc` (Stan transpiler) option `warn-pedantic`. Here we
demonstrate pedantic mode with a Stan program that is syntactically correct but
is missing a lower bound and a prior for a parameter.

```{r stan_file_pedantic}
stan_file_pedantic <- write_stan_file("
data {
  int N;
  array[N] int y;
}
parameters {
  // should have <lower=0> but omitting to demonstrate pedantic mode
  real lambda;
}
model {
  y ~ poisson(lambda);
}
")
```

To turn on pedantic mode when compiling set `pedantic=TRUE` in the call to
`cmdstan_model()`.

```{r pedantic-compile, collapse = TRUE}
mod_pedantic <- cmdstan_model(stan_file_pedantic, pedantic = TRUE)
```

To turn on pedantic mode separately from compilation use the `pedantic` argument
to the `$check_syntax()` method.

```{r pedantic-check_syntax, collapse=TRUE}
mod_pedantic$check_syntax(pedantic = TRUE)
```

To check a program without compiling it, which is much faster, use
`check_syntax_stan_file()` on the file itself.

```{r pedantic-check_syntax-2, collapse=TRUE}
check_syntax_stan_file(stan_file_pedantic, pedantic = TRUE)
```

### Stan model variables

You can obtain the names, types and dimensions of the data, parameters,
transformed parameters and generated quantities variables of a Stan model using
the `$variables()` method of the `CmdStanModel` object, or with
`variables_stan_file()` on the file itself when there is no need to compile the
program, as in the case below.

```{r variables_stan_file}
variables_stan_file <- write_stan_file("
data {
  int<lower=1> J;
  vector<lower=0>[J] sigma;
  vector[J] y;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] theta_raw;
}
transformed parameters {
  vector[J] theta = mu + tau * theta_raw;
}
model {
  target += normal_lpdf(tau | 0, 10);
  target += normal_lpdf(mu | 0, 10);
  target += normal_lpdf(theta_raw | 0, 1);
  target += normal_lpdf(y | theta, sigma);
}
")
variables <- variables_stan_file(variables_stan_file)
```

The result is a list with `data`, `parameters`,
`transformed_parameters` and `generated_quantities` elements, each
corresponding to variables in their respective block of the program. Transformed
data variables are not listed as they are not used in the model's input
or output.

```{r variables-list-names}
names(variables)
names(variables$data)
names(variables$parameters)
names(variables$transformed_parameters)
names(variables$generated_quantities)
```

Each variable is represented as a list containing the type information
(`real`, `int` or `complex`, or for a tuple a list of its elements' entries)
and the number of dimensions. Here "number of dimensions" means the number of
indexing dimensions in the declared Stan variable (scalar = 0, vector = 1,
matrix = 2, etc.). See `?variables` for more details.

```{r variable-type-dims}
variables$data$J
variables$data$sigma
variables$parameters$tau
variables$transformed_parameters$theta
```

### Executable location

By default, the executable is created in the same directory as the file
containing the Stan program. You can also specify a different location with the
`dir` argument.

```{r compile-with-dir, eval = FALSE}
mod <- cmdstan_model(stan_file, dir = "path/to/directory/for/executable")
```

### The build record

Starting with v1.0, to help it determine whether a rebuild is necessary CmdStanR
writes a hidden JSON file next to the executable that records how the executable
was built: the Stan file and the files it includes, the options passed to
`stanc` and `make`, the CmdStan installation, and a hash of the executable
itself. The file is named after the executable, so `bernoulli` is described by
`.bernoulli.cmdstanr.json`.

```{r build-record}
list.files(
  dirname(mod$exe_file()), pattern = "bernoulli.*cmdstanr.json", all.files = TRUE
)
```

The `$build_info()` method reads the record back. The standalone
`stan_build_info()` function does the same for a path to an executable when you
don't have a model object. When there is no usable record it says why and
reports only what the executable says about itself.

```{r build-info}
mod$build_info()

# demonstrating stan_build_info even though here we have the model object 
# so we can use mod$build_info()
stan_build_info(mod$exe_file())
```

If the directory is under version control, we recommend adding the record file
beside whatever already excludes the executable in `.gitignore` (the leading dot
doesn't hide the file from git). You can ignore all record files using
`.*.cmdstanr.json` or ignore them individually by name.

If the directory is inside an R package, we recommend adding a matching regular
expression to `.Rbuildignore`, such as `\..*\.cmdstanr\.json$`.

In general, if you move the executable somewhere move the build record with it.
If a build step copies the executable somewhere else, such as a CI artifact, a
container image or a shared build directory, copy the record with it.

### What the build record does not track

`cmdstan_model()` rebuilds when the record no longer matches the Stan program,
the files it includes, the user header, the build options, `make/local` or the
CmdStan installation. Some things that change a build are not in the record, so
changing them does not automatically trigger a rebuild and you should manually
force recompilation:

- the C++ toolchain
- environment variables the compiler or `make` reads
- CmdStan or the Stan Math library modified in place
- a makefile that `make/local` includes
- headers that a `user_header` includes

For the last two, the build prints a one-line note when it sees the `include`
line, since it can tell that there is something it does not track. The note
only covers what the scan can see. A header of your own that the user header
includes with angle brackets, as in `#include <helpers.hpp>` with its
directory added through `cpp_options`, is not noticed, because the scan
cannot tell one of your headers from a standard one like `<vector>` without
searching the include path. No note means nothing was detected, not that
every dependency is tracked.

When one of these has changed, pass `force_recompile = TRUE` to
`cmdstan_model()` or set `options(cmdstanr_force_recompile = TRUE)` for the
session.

## Processing data

There are three data formats that CmdStanR allows when fitting a model:

* Named list of R objects
* JSON file
* R dump file

### Named list of R objects

Like the RStan interface, CmdStanR accepts a named list of R objects where the
names correspond to variables declared in the data block of the Stan program.
In the Bernoulli model the data is `N`, the number of data points, and `y`
an integer array of observations.

```{r print-program-again}
mod$print()
```

```{r data-list, eval=FALSE}
# data block has 'N' and 'y'
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list)
```

Because CmdStan doesn't accept lists of R objects, CmdStanR will first write the
data to a temporary JSON file using `write_stan_json()`. This happens
internally, but it is also possible to call `write_stan_json()` directly.

```{r write_stan_json}
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)
cat(readLines(json_file), sep = "\n")
```

### JSON file

If you already have your data in a JSON file you can just pass that file
directly to CmdStanR instead of using a list of R objects. For example, we could
pass in the JSON file we created above using `write_stan_json()`:

```{r data-json, eval=FALSE}
fit <- mod$sample(data = json_file)
```

### R dump file

CmdStanR also accepts data in the R dump format because CmdStan supports it.
However, JSON is faster for CmdStan to process and is the recommended file
format.


## Writing CmdStan output to CSV

### Default temporary files

```{r sample-tempdir, results = "hide"}
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list)
```

When fitting a model, the default behavior is to write the output from CmdStan
to CSV files in a temporary directory.

```{r output_files}
fit$output_files()
```

These files will be lost if you end your R session or if you remove the
`fit` object and force (or wait for) garbage collection.


```{r gc}
files <- fit$output_files()
file.exists(files)

rm(fit)
gc()

file.exists(files)
```

### Non-temporary files

To save these files to a non-temporary location there are two options. You
can either specify the `output_dir` argument to `mod$sample()` or use
`fit$save_output_files()` after fitting the model.

```{r save_output_files, eval=FALSE}
# see ?save_output_files for info on optional arguments
fit$save_output_files(dir = "path/to/directory")
```

```{r output_dir, eval = FALSE}
fit <- mod$sample(
  data = data_list,
  output_dir = "path/to/directory"
)
```


## Reading CmdStan output into R

### Lazy CSV reading

With the exception of some diagnostic information, the CSV files are not read
into R until their contents are requested by calling a method that requires them
(e.g., `fit$draws()`, `fit$summary()`, etc.). If we examine the structure of the
`fit` object, notice how the `Private` slot `draws_` is `NULL`, indicating that
the CSV files haven't yet been read into R.

```{r refit, include=FALSE}
fit <- mod$sample(data = data_list)
```
```{r csv-not-read}
str(fit)
```

After we call a method that requires the draws then if we reexamine the
structure of the object we will see that the `draws_` slot in `Private`
is no longer empty.

```{r for-csv-reading}
draws <- fit$draws() # force CSVs to be read into R
str(fit)
```

For models with many parameters, transformed parameters, or generated
quantities, if only some are requested (e.g., by specifying the `variables`
argument to `fit$draws()`) then CmdStanR will only read in the requested
variables (unless they have already been read in).

### read_cmdstan_csv()

Internally, the `read_cmdstan_csv()` function is used to read the CmdStan CSV
files into R. This function is exposed to users, so you can also call it
directly.

```{r read_cmdstan_csv}
# see ?read_cmdstan_csv for info on optional arguments controlling
# what information is read in
csv_contents <- read_cmdstan_csv(fit$output_files())
str(csv_contents)
```

### as_cmdstan_fit()

If you need to manually create fitted model objects from CmdStan CSV files use
`as_cmdstan_fit()`.

```{r as_cmdstan_fit}
fit2 <- as_cmdstan_fit(fit$output_files())
```

This is pointless in our case since we have the original `fit` object, but this
function can be used to create fitted model objects (`CmdStanMCMC`,
`CmdStanMLE`, etc.) from CmdStan CSV files.

### Saving and accessing advanced algorithm info (latent dynamics)

If `save_latent_dynamics` is set to `TRUE` when running the `$sample()` method
then additional CSV files are created (one per chain) that provide access to
quantities used under the hood by Stan's implementation of dynamic Hamiltonian
Monte Carlo.

CmdStanR does not yet provide a special method for processing these files but
they can be read into R using R's standard CSV reading functions.


```{r save_latent_dynamics, results = "hide"}
fit <- mod$sample(data = data_list, save_latent_dynamics = TRUE)
```
```{r read-latent-dynamics}
fit$latent_dynamics_files()

# read one of the files in
x <- utils::read.csv(fit$latent_dynamics_files()[1], comment.char = "#")
head(x)
```

The column `lp__` is also provided via `fit$draws()`, and the columns
`accept_stat__`, `stepsize__`, `treedepth__`, `n_leapfrog__`, `divergent__`, and
`energy__` are also provided by `fit$sampler_diagnostics()`, but there are
several columns unique to the latent dynamics file.

```{r explore-latent-dynamics}
head(x[, c("theta", "p_theta", "g_theta")])
```

Our model has a single parameter `theta` and the three columns above correspond
to `theta` in the _unconstrained_ space (`theta` on the constrained space is
accessed via `fit$draws()`), the auxiliary momentum `p_theta`, and the gradient
`g_theta`. In general, these three quantities are recorded for every parameter
in the model. Container parameters are organized into columns for each scalar
element in the unconstrained parameterization. CmdStan uses period-separated
indices for these elements, for example `theta.1`, `p_theta.1`, and `g_theta.1`.
See the [diagnostic CSV output documentation](https://mc-stan.org/docs/cmdstan-guide/stan_csv_apdx.html#diagnostic-csv-output-file) for more details.


## Developing using CmdStanR

CmdStanR can of course be used for developing other packages that require
compiling and running Stan models as well as using new or custom Stan features
available through CmdStan. Even though CmdStanR is not itself on CRAN, it
can still be used by packages that are on CRAN.

### Compiling Stan models during package installation

Stan models used by an R package can be compiled at runtime or during
installation from source. Installation-time compilation builds executables on
the installing system so they can be reused without recompilation at runtime.
This is different from including compiled executables in the source package:
compiled artifacts are generally not portable across platforms, and
[CRAN source packages may not contain binary executable code](https://cran.r-project.org/web/packages/policies.html#Source-packages).

The [`instantiate`](https://wlandau.github.io/instantiate/) package provides a
current workflow for installation-time compilation. See the `instantiate`
documentation for requirements, limitations, and advice on preparing for CRAN
submission.


### Troubleshooting and debugging

When developing or testing new features it might be useful to have more
information on how CmdStan is called internally and to see more information
printed when compiling or running models. This can be enabled for an entire R
session by setting the option `"cmdstanr_verbose"` to `TRUE`.

```{r include=FALSE}
old_options <- options("cmdstanr_verbose")
```
```{r verbose-mode}
options(cmdstanr_verbose = TRUE)

mod <- cmdstan_model(stan_file, force_recompile = TRUE)
fit <- mod$sample(
  data = data_list,
  chains = 1,
  iter_warmup = 100,
  iter_sampling = 100
)
```
```{r include=FALSE}
options(old_options)
```
