---
title: "Getting started with CmdStanR"
author: "Jonah Gabry"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Getting started with CmdStanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r settings-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  # collapse = TRUE,
  comment=NA,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```


```{r setup, message=FALSE}
library(cmdstanr)
```


## Introduction

CmdStanR is a lightweight interface to [Stan](https://mc-stan.org) for R users
(see [CmdStanPy](https://github.com/stan-dev/cmdstanpy) for Python) that
provides an alternative to the traditional [RStan](https://mc-stan.org/rstan)
interface. See the *Comparison with RStan* section later in this vignette
for more details on how the two inferfaces differ.

**CmdStanR is unreleased and user facing functions are still subject to change**

## Installing CmdStan

CmdStanR requires a working installation of
[CmdStan](https://mc-stan.org/users/interfaces/cmdstan), the shell interface to
Stan. If you don't have CmdStan installed you can install it yourself or use the
`install_cmdstan()` function provided by CmdStanR.

```{r, include = FALSE}
if (!dir.exists(cmdstan_default_path())) {
  install_cmdstan()
}
```
```{r, eval=FALSE}
install_cmdstan()
```

Before CmdStanR can be used it needs to know where the CmdStan installation is
located. When the package is loaded it tries to help automate this to avoid
having to manually set the path every session:

1. If the environment variable `"CMDSTAN"` exists at load time then its value will be automatically set as the default path to CmdStan for the R session. This is useful
if your CmdStan installation is not located in the default directory that would
have been used by `install_cmdstan()` (see #2).

2. If no environment variable is found when loaded but the directory `".cmdstanr/cmdstan"` exists in the user's *home* directory (not working directory) then it will be set as the path to CmdStan for the R session. This is the same as the default directory that `install_cmdstan()` uses to install the latest version of CmdStan, so if that's how you installed CmdStan you shouldn't need to manually 
set the path to CmdStan when loading CmdStanR.

If neither of these applies (or you want to subsequently change the path) you
can use the `set_cmdstan_path()` function:

```{r set_cmdstan_path, eval=FALSE}
set_cmdstan_path(PATH_TO_CMDSTAN)
```

If you need to check or get the current path to CmdStan use the 
[`cmdstan_path()`](https://mc-stan.org/cmdstanr/reference/cmdstan_path.html)
function.

## Compiling and running Stan's algorithms

### Compilation 

The `cmdstan_model()` function creates a new
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html)
object from a file containing a Stan program. Here we'll use the example Stan
program that comes with the CmdStan installation:

```{r cmdstan_model}
file <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli.stan")
mod <- cmdstan_model(file)
```

The object `mod` is an [R6](https://r6.r-lib.org/) object, which is similar to 
R's reference class objects. This design choice was made to make CmdStanR
and CmdStanPy have similar user interfaces. 

The
[`$compile()`](https://mc-stan.org/cmdstanr/reference/model-method-compile.html)
method is used to translate the Stan code to C++ and compile the C++ code:

```{r compile}
mod$compile()
mod$print()  # print the Stan program
```


### Model fitting

To run Stan's MCMC algorithm we can use the
[`$sample()`](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)
method for
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html)
objects. Data can be passed in as a named list of R objects (like for RStan) or
as a path to a data file compatible with CmdStan (R dump or JSON).

```{r sample}
data_list <- list(N = 10, y =c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list, seed = 123, num_chains = 2)
```

```{r summary}
# calls CmdStan's bin/stansummary
fit$summary()
```

#### Accessing the contents of fitted model objects

CmdStanR is still unreleased and not fully developed. Currently the best way
to work with the output is to create a stanfit object (RStan) from the 
csv CmdStan output files:

```{r stanfit}
stanfit <- rstan::read_stan_csv(fit$output_files())
print(stanfit)
```

## Comparison with RStan

The RStan interface ([**rstan** package](https://mc-stan.org/rstan) package) is
an in-memory interface to Stan and relies on R packages like **Rcpp** and
**inline** call C++ code from R. On the other hand, the CmdStanR interface does
not directly call any C++ code from R, instead relying on CmdStan for
compilation, running algorithms, and writing results to output files.

Both forms of interfacing with Stan have advantages and disadvantages. An
in-memory interface like RStan is able to offer more advanced features than
CmdStanR (for example RStan's `grad_log_prob` and `unconstrain_pars` methods)
but keeping up with Stan releases is more complicated for RStan, often requiring
non-trivial changes to the **rstan** package and requiring new CRAN releases of
**rstan** and **StanHeaders**. With CmdStanR, the latest features in Stan will
be available from R immediately after updating CmdStan, without an update to
the **cmdstanr** package.

