---
title: "How does CmdStanR work?"
author: "Jonah Gabry and Rok Češnovar"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{How does CmdStanR work?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="children/settings-knitr.Rmd"}
```

## Introduction

This vignette is intended to be read after the [_Getting started with CmdStanR_](http://mc-stan.org/cmdstanr/articles/cmdstanr.html) 
vignette. Please read that first for important background. In this document we
provide additional details about compiling models, passing in data, and how
CmdStan output is saved and read back into R.

We will only use the `$sample()` method in examples, but all model fitting
methods work in a similar way under the hood.

```{r setup, message=FALSE}
library(cmdstanr)
```

## Compilation

### Immediate compilation

The `cmdstan_model()` function creates a new `CmdStanModel` object. The
`CmdStanModel` object stores the path to a Stan program as well as the 
path to a compiled executable: 

```{r start-clean, include=FALSE}
exe <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli")
unlink(exe)
```

```{r compile}
stan_file <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli.stan")
mod <- cmdstan_model(stan_file)
mod$print()
mod$stan_file()
mod$exe_file()
```

Subsequently, if you create a `CmdStanModel` object from the same Stan file
then compilation will be skipped (assuming the file hasn't changed):

```{r already-compiled}
mod <- cmdstan_model(stan_file)
```
 
Internally, `cmdstan_model()` first creates the `CmdStanModel` object from
just the Stan file and then calls its [`$compile()`](http://mc-stan.org/cmdstanr/reference/model-method-compile.html) 
method. Optional arguments to the `$compile()` method can be passed via `...`, 
for example:

```{r compile-options, eval=FALSE}
mod <- cmdstan_model(
  stan_file, 
  force_recompile = TRUE, 
  include_paths = "paths/to/directories/with/included/files", 
  cpp_options = list(stan_threads = TRUE, STANC2 = TRUE)
)
```



### Delayed compilation

It is also possible to delay compilation when creating the `CmdStanModel` object
by specifying `compile=FALSE`. You can later call the `$compile()` method
directly:

```{r compile-method}
unlink(mod$exe_file())
mod <- cmdstan_model(stan_file, compile = FALSE)
mod$exe_file() # not yet created
mod$compile()
mod$exe_file()
```

### Pedantic check

If you are using CmdStan version 2.24 or later and CmdStanR version 0.2.1 or
later, you can run a pedantic check for your model. CmdStanR will always check
that your Stan program does not contain any invalid syntax but with pedantic
mode enabled the check will also warn you about other potential issues in your
model, for example:

- Distribution usages issues: distribution arguments do not match the
distribution specification, or some specific distribution is used in an
inadvisable way.
- Unused parameter: a parameter is defined but does not contribute to target.
- Large or small constant in a distribution: very large or very small constants
are used as distribution arguments.
- Control flow depends on a parameter: branching control flow (like if/else)
depends on a parameter value.
- Parameter has multiple twiddles: a parameter is on the left-hand side of
multiple twiddles (i.e., multiple `~` symbols).
- Parameter has zero or multiple priors: a parameter has zero or more than one
prior distribution.
- Variable is used before assignment: a variable is used before being assigned a
value.
- Strict or nonsensical parameter bounds: a parameter is given questionable
bounds.

Pedantic mode is available when compiling the model or when using the separate
`$check_syntax()` method of a `CmdStanModel` object. Internally this corresponds
to setting the `stanc` (Stan transpiler) option `warn-pedantic`. Here we
demonstrate pedantic mode with a Stan program that is syntactically correct but
is missing a lower bound and a prior for a parameter.

```{r stan_file_pedantic}
stan_file_pedantic <- write_stan_file("
data {
  int N;
  int y[N];
}
parameters {
  // should have <lower=0> but omitting to demonstrate pedantic mode
  real lambda;
}
model {
  y ~ poisson(lambda);
}
")
```

To turn on pedantic mode at compile time you can set `pedantic=TRUE` in 
the call to `cmdstan_model()` (or when calling the `$compile()` method directly
if using the delayed compilation approach described above).

```{r pedantic-compile, collapse = TRUE}
mod_pedantic <- cmdstan_model(stan_file_pedantic, pedantic = TRUE)
```

To turn on pedantic mode separately from compilation use the `pedantic` argument
to the `$check_syntax()` method.

```{r pedantic-check_syntax, collapse=TRUE}
mod_pedantic$check_syntax(pedantic = TRUE) 
```

Using `pedantic=TRUE` via the `$check_syntax()` method also has the advantage
that it can be used even if the model hasn't been compiled yet. This can be
helpful because the pedantic and syntax checks themselves are much faster than
compilation.

```{r pedantic-check_syntax-2, collapse=TRUE}
file.remove(mod_pedantic$exe_file()) # delete compiled executable
rm(mod_pedantic)

mod_pedantic <- cmdstan_model(stan_file_pedantic, compile = FALSE)
mod_pedantic$check_syntax(pedantic = TRUE)
```

### Stan model variables

If using CmdStan 2.27 or newer, you can obtain the names, types
and dimensions of the data, parameters, transformed parameters
and generated quantities variables of a Stan model using the
`$variables()` method of the `CmdStanModel` object.

```{r stan_file_variables}
stan_file_variables <- write_stan_file("
data {
  int<lower=1> J;
  vector<lower=0>[J] sigma;
  vector[J] y;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] theta_raw;
}
transformed parameters {
  vector[J] theta = mu + tau * theta_raw;
}
model {
  target += normal_lpdf(tau | 0, 10);
  target += normal_lpdf(mu | 0, 10);
  target += normal_lpdf(theta_raw | 0, 1);
  target += normal_lpdf(y | theta, sigma);
}
")
mod_v <- cmdstan_model(stan_file_variables)
variables <- mod_v$variables()
```

The `$variables()` method returns a list with `data`, `parameters`,
`transformed_parameters` and `generated_quantities` elements, each
corresponding to variables in their respective block of the program. Transformed 
data variables are not listed as they are not used in the model's input
or output.

```{r variables-list-names}
names(variables)
names(variables$data)
names(variables$parameters)
names(variables$transformed_parameters)
names(variables$generated_quantities)
```

Each variable is represented as a list containing the type
information (currently limited to `real` or `int`) and the number of dimensions.

```{r variable-type-dims}
variables$data$J
variables$data$sigma
variables$parameters$tau
variables$transformed_parameters$theta
```

### Executable location

By default, the executable is created in the same directory as the file
containing the Stan program. You can also specify a different location with the
`dir` argument:

```{r compile-with-dir, eval = FALSE}
mod <- cmdstan_model(stan_file, dir = "path/to/directory/for/executable")
```

## Processing data

There are three data formats that CmdStanR allows when fitting a model:

* named list of R objects 
* JSON file
* R dump file

### Named list of R objects

Like the RStan interface, CmdStanR accepts a named list of R objects where the
names correspond to variables declared in the data block of the Stan program. 
In the Bernoulli model the data is `N`, the number of data points, and `y`
an integer array of observations. 

```{r print-program-again}
mod$print()
```

```{r data-list, eval=FALSE}
# data block has 'N' and 'y'
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list)
```

Because CmdStan doesn't accept lists of R objects, CmdStanR will first write the
data to a temporary JSON file using `write_stan_json()`. This happens
internally, but it is also possible to call `write_stan_json()` directly:

```{r write_stan_json}
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)
cat(readLines(json_file), sep = "\n")
```

### JSON file

If you already have your data in a JSON file you can just pass that file
directly to CmdStanR instead of using a list of R objects. For example, we could
pass in the JSON file we created above using `write_stan_json()`:

```{r data-json, eval=FALSE}
fit <- mod$sample(data = json_file)
```


### R dump file

Finally, it is also possible to use the R dump file format. This is *not*
recommended because CmdStan can process JSON faster than R dump, but CmdStanR
allows it because CmdStan will accept files created by `rstan::stan_rdump()`:

```{r data-rdump, eval=FALSE}
rdump_file <- tempfile(fileext = ".data.R")
rstan::stan_rdump(names(data_list), file = rdump_file, envir = list2env(data_list))
cat(readLines(rdump_file), sep = "\n")
fit <- mod$sample(data = rdump_file)
```


## Writing CmdStan output to CSV

### Default temporary files

```{r sample-tempdir, results = "hide"}
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list)
```

When fitting a model, the default behavior is to write the output from CmdStan 
to CSV files in a temporary directory:

```{r output_files}
fit$output_files()
```

These files will be lost if you end your R session or if you remove the 
`fit` object and force (or wait for) garbage collection. 


```{r gc}
files <- fit$output_files()
file.exists(files)

rm(fit)
gc()

file.exists(files)
```

### Non-temporary files

To save these files to a non-temporary location there are two options. You 
can either specify the `output_dir` argument to `mod$sample()` or use 
`fit$save_output_files()` after fitting the model: 


```{r save_output_files, eval=FALSE}
# see ?save_output_files for info on optional arguments
fit$save_output_files(dir = "path/to/directory")
```

```{r output_dir, eval = FALSE}
fit <- mod$sample(
  data = data_list, 
  output_dir = "path/to/directory"
)
```


## Reading CmdStan output into R

### Lazy CSV reading 

With the exception of some diagnostic information, the CSV files are not read
into R until their contents are requested by calling a method that requires them
(e.g., `fit$draws()`, `fit$summary()`, etc.). If we examine the structure of the
`fit` object, notice how the `Private` slot `draws_` is `NULL`, indicating that
the CSV files haven't yet been read into R:

```{r refit, include=FALSE}
fit <- mod$sample(data = data_list)
```
```{r csv-not-read}
str(fit)
```

After we call a method that requires the draws then if we reexamine the 
structure of the object we will see that the `draws_` slot in `Private` 
is no longer empty: 

```{r for-csv-reading}
draws <- fit$draws() # force CSVs to be read into R
str(fit)
```

For models with many parameters, transformed parameters, or generated
quantities, if only some are requested (e.g., by specifying the `variables`
argument to `fit$draws()`) then CmdStanR will only read in the requested
variables (unless they have already been read in).

### read_cmdstan_csv()

Internally, the `read_cmdstan_csv()` function is used to read the CmdStan CSV
files into R. This function is exposed to users, so you can also call it
directly:

```{r read_cmdstan_csv}
# see ?read_cmdstan_csv for info on optional arguments controlling 
# what information is read in
csv_contents <- read_cmdstan_csv(fit$output_files())
str(csv_contents)
```


### Saving and accessing advanced algorithm info (latent dynamics)

If `save_latent_dynamics` is set to `TRUE` when running the `$sample()` method
then additional CSV files are created (one per chain) that provide access to
quantities used under the hood by Stan's implementation of dynamic Hamiltonian
Monte Carlo.

CmdStanR does not yet provide a special method for processing these files but
they can be read into R using R's standard CSV reading functions:


```{r save_latent_dynamics, results = "hide"}
fit <- mod$sample(data = data_list, save_latent_dynamics = TRUE)
```
```{r read-latent-dynamics}
fit$latent_dynamics_files()

# read one of the files in
x <- utils::read.csv(fit$latent_dynamics_files()[1], comment.char = "#")
head(x)
```

The column `lp__` is also provided via `fit$draws()`, and the columns
`accept_stat__`, `stepsize__`, `treedepth__`, `n_leapfrog__`, `divergent__`, and
`energy__` are also provided by `fit$sampler_diagnostics()`, but there are 
several columns unique to the latent dynamics file:

```{r explore-latent-dynamics}
head(x[, c("theta", "p_theta", "g_theta")])
```

Our model has a single parameter `theta` and the three columns above correspond
to `theta` in the _unconstrained_ space (`theta` on the constrained space is
accessed via `fit$draws()`), the auxiliary momentum `p_theta`, and the gradient
`g_theta`. In general, each of these three columns will exist for _every_
parameter in the model.


## Saving fitted model objects

As described above, the contents of the CSV files are only read into R when they
are needed. This means that in order to save a fitted model object containing
_all_ of the posterior draws and sampler diagnostics you should either make sure
to call `fit$draws()` and `fit$sampler_diagnostics()` before saving the object
`fit`, or use the special `$save_object()` method provided by CmdStanR, which
will ensure that everything has been read into R before saving the object using
`saveRDS()`:

```{r save-fit}
temp_rds_file <- tempfile(fileext = ".RDS") # temporary file just for demonstration
fit$save_object(file = temp_rds_file)
```

We can check that this worked by removing `fit` and loading it back in from
the save file:

```{r load-fit}
rm(fit); gc()

fit <- readRDS(temp_rds_file)
fit$summary()
```

## Developing using CmdStanR

CmdStanR can of course be used for developing other packages that require compiling
and running Stan models as well as using new or custom Stan features available
through CmdStan.

### Troubleshooting and debugging

When developing or testing new features it might be useful to have more
information on how CmdStan is called internally and to see more information
printed when compiling or running models. This can be enabled for an entire R
session by setting the option `"cmdstanr_verbose"` to `TRUE`.

```{r verbose-mode}
options("cmdstanr_verbose"=TRUE)

mod <- cmdstan_model(stan_file, force_recompile = TRUE)
fit <- mod$sample(
  data = data_list,
  chains = 1,
  iter_warmup = 100,
  iter_sampling = 100
)
```

```{r include=FALSE}
options("cmdstanr_verbose" = FALSE)
```
