---
title: "Getting started with CmdStanR"
author: "Jonah Gabry and Rok Češnovar"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Getting started with CmdStanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r settings-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  # collapse = TRUE,
  dev = "png",
  dpi = 150,
  fig.asp = 0.618,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center",
  comment = NA,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```


## Introduction

CmdStanR is a lightweight interface to [Stan](https://mc-stan.org) for R users
(see [CmdStanPy](https://github.com/stan-dev/cmdstanpy) for Python) that
provides an alternative to the traditional [RStan](https://mc-stan.org/rstan)
interface. See the [*Comparison with RStan*](#comparison-with-rstan) section
later in this vignette for more details on how the two interfaces differ.

**CmdStanR is not on CRAN yet**, but the latest pre-beta release can be
installed via:

```{r install, eval=FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

CmdStanR (the **cmdstanr** R package) can now be loaded like any other R package. 
We'll also load the **bayesplot** and **posterior** packages.

```{r library, message=FALSE}
library(cmdstanr)
library(posterior)
library(bayesplot)
color_scheme_set("brightblue")
```

## Installing CmdStan

CmdStanR requires a working installation of
[CmdStan](https://mc-stan.org/users/interfaces/cmdstan), the shell interface to
Stan. If you don't have CmdStan installed you can install it yourself or use the
`install_cmdstan()` function provided by CmdStanR.

```{r, include = FALSE}
if (!dir.exists(cmdstan_default_path())) {
  install_cmdstan()
}
```
```{r, eval=FALSE}
install_cmdstan(cores = 2)
```

Before CmdStanR can be used it needs to know where the CmdStan installation is
located. When the package is loaded it tries to help automate this to avoid
having to manually set the path every session:

1. If the environment variable `"CMDSTAN"` exists at load time then its value
will be automatically set as the default path to CmdStan for the R session. This
is useful if your CmdStan installation is not located in the default directory
that would have been used by `install_cmdstan()` (see #2).

2. If no environment variable is found when loaded but any directory in the form
`".cmdstanr/cmdstan-[version]"`, for example `".cmdstanr/cmdstan-2.23.0"`,
exists in the user's home directory (`Sys.getenv("HOME")`,
*not* the current working directory) then the path to the cmdstan with the
largest version number will be set as the path to CmdStan for the R session.
This is the same as the default directory that `install_cmdstan()` uses to
install the latest version of CmdStan, so if that's how you installed CmdStan
you shouldn't need to manually set the path to CmdStan when loading CmdStanR.

If neither of these applies (or you want to subsequently change the path) you
can use the `set_cmdstan_path()` function:

```{r set_cmdstan_path, eval=FALSE}
set_cmdstan_path(PATH_TO_CMDSTAN)
```

To check the path to the CmdStan installation and the CmdStan version number
you can use `cmdstan_path()` and `cmdstan_version()`:

```{r cmdstan_path}
cmdstan_path()
cmdstan_version()
```

## Compiling a model

The `cmdstan_model()` function creates a new
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html)
object from a file containing a Stan program. Under the hood, CmdStan is called
to translate a Stan program to C++ and create a compiled executable. Here we'll
use the example Stan program that comes with the CmdStan installation:

```{r cmdstan_model}
file <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli.stan")
mod <- cmdstan_model(file)
```

The object `mod` is an [R6](https://r6.r-lib.org/) reference object of class
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html) and
behaves similarly to R's reference class objects and those in object oriented
programming languages. Methods are accessed using the `$` operator. This design
choice allows for CmdStanR and
[CmdStanPy](https://github.com/stan-dev/cmdstanpy) to provide a similar user
experience and share many implementation details.

The Stan program can be printed using the `$print()` method: 

```{r compile}
mod$print()
```

The path to the compiled executable is returned by the `$exe_file()`
method:

```{r exe_file}
mod$exe_file()
```

## Fitting a model

### MCMC

The
[`$sample()`](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)
method for
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html)
objects runs Stan's default MCMC algorithm. The `data` argument accepts a named
list of R objects (like for RStan) or a path to a data file compatible with
CmdStan (JSON or R dump).

```{r sample}
# names correspond to the data block in the Stan program
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))

fit <- mod$sample(
  data = data_list, 
  seed = 123, 
  chains = 4, 
  parallel_chains = 2,
  refresh = 500
)
```

There are many more arguments that can be passed to the `$sample()` method. 
For details follow this link to its separate documentation page: 

* [`$sample()`](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)

The `$sample()` method creates [R6](https://r6.r-lib.org/) `CmdStanMCMC`
objects, which have many associated methods. Below we will demonstrate some of
the most important methods. For a full list, follow this link to the
`CmdStanMCMC` documentation:

* [`CmdStanMCMC`](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html)

#### Posterior summary statistics

The
[`$summary()`](https://mc-stan.org/cmdstanr/reference/fit-method-summary.html)
method calls `summarise_draws()` from the **posterior** package:

```{r summary}
fit$summary()
fit$summary("theta", "mean", "sd")

# use a formula to summarize arbitrary functions, e.g. Pr(theta <= 0.5)
fit$summary("theta", pr_lt_half = ~ mean(. <= 0.5))
```


#### Posterior draws

The [`$draws()`](https://mc-stan.org/cmdstanr/reference/fit-method-draws.html)
method can be used to extract the posterior draws as a 3-D array (iteration x
chain x variable). The **posterior** package can then be used to easily convert
to other formats, like data frames and matrices of draws.

```{r draws, message=FALSE}
# this is a draws_array object from the posterior package
draws_array <- fit$draws()
str(draws_array)

# convert to matrix or data frame 
draws_df <- as_draws_df(draws_array) # as_draws_matrix() for matrix
print(draws_df)
```

Plotting posterior distributions is as easy as passing the object returned by
the `$draws()` method directly to plotting functions in
[**bayesplot**](https://mc-stan.org/bayesplot):

```{r plots, message=FALSE}
mcmc_hist(fit$draws("theta"))
```

#### Sampler diagnostics

The
[`$sampler_diagnostics()`](https://mc-stan.org/cmdstanr/reference/fit-method-sampler_diagnostics.html)
method extracts the values of the sampler parameters (`treedepth__`,
`divergent__`, etc.) as a 3-D array (iteration x chain x variable):

```{r sampler_diagnostics}
# this is a draws_array object from the posterior package
str(fit$sampler_diagnostics())

# convert to matrix or data frame using posterior package
diagnostics_df <- as_draws_df(fit$sampler_diagnostics())
print(diagnostics_df)
```

#### CmdStan utilities

The [`$cmdstan_diagnose()`](https://mc-stan.org/cmdstanr/reference/fit-method-cmdstan_summary.html)
and [`$cmdstan_summary()`](https://mc-stan.org/cmdstanr/reference/fit-method-cmdstan_summary.html)
methods call CmdStan's `diagnose` and `stansummary` utilities:

```{r summary-and-diagnose}
fit$cmdstan_diagnose()
fit$cmdstan_summary()
```


#### Create a `stanfit` object

If you have RStan installed then it is also possible to create a `stanfit` 
object from the csv output files written by CmdStan. This can be done by using
`rstan::read_stan_csv()` in combination with the `$output_files()` method of the
`CmdStanMCMC` object:

```{r stanfit, eval=FALSE}
stanfit <- rstan::read_stan_csv(fit$output_files())
```



### Optimization and variational inference

CmdStanR also supports running Stan's optimization algorithms and its algorithms
for variational approximation of full Bayesian inference. These are run via the
`$optimize()` and `$variational()` methods, which are called in a similar way to
the `$sample()` method demonstrated above.

#### Optimization

We can find the (penalized) maximum likelihood estimate (MLE) using [`$optimize()`](https://mc-stan.org/cmdstanr/reference/model-method-optimize.html):

```{r optimize}
fit_mle <- mod$optimize(data = data_list, seed = 123) 
fit_mle$summary() # includes lp__ (log prob calculated by Stan program)
fit_mle$mle("theta") 
```

Here's a plot comparing the MLE to the posterior distribution of `theta`:

```{r plot-mle, message = FALSE}
mcmc_hist(fit$draws("theta")) + 
  vline_at(fit_mle$mle(), size = 1.5)
```

#### Variational Bayes

We can run Stan's experimental variational Bayes algorithm using the 
[`$variational()`](https://mc-stan.org/cmdstanr/reference/model-method-variational.html) 
method:

```{r variational}
fit_vb <- mod$variational(data = data_list, seed = 123, output_samples = 4000) 
fit_vb$summary("theta")
```
The `$draws()` method can be used to access the approximate posterior draws.
Let's extract the draws, make the same plot we made after MCMC, and compare the
two. In this trivial example the distributions look quite similar, although
the variational approximation slightly underestimates the posterior 
standard deviation:

```{r plot-variational, message = FALSE}
bayesplot_grid(
  mcmc_hist(fit$draws("theta"), binwidth = 0.025),    
  mcmc_hist(fit_vb$draws("theta"), binwidth = 0.025),
  titles = c("Posterior distribution from MCMC", "Approximate posterior from VB"),
  xlim = c(0, 1)
)
```


For more details on the `$optimize()` and `$variational()` methods, follow
these thinks to their documentation pages:

* [`$optimize()`](https://mc-stan.org/cmdstanr/reference/model-method-optimize.html)
* [`$variational()`](https://mc-stan.org/cmdstanr/reference/model-method-variational.html)


## Saving fitted model objects

In order to save a fitted model object to disk and ensure that all of the
contents are available when reading the object back into R, we recommend using the
[`$save_object()`](http://mc-stan.org/cmdstanr/reference/fit-method-save_object.html)
method provided by CmdStanR. The reason for this is discussed in detail in the vignette
[_How does CmdStanR work?_](http://mc-stan.org/cmdstanr/articles/cmdstanr-internals.html), 
so here we just demonstrate how to use the method:

```{r save_object, eval=FALSE}
fit$save_object(file = "fit.RDS")

# can be read back in using readRDS
fit2 <- readRDS("fit.RDS")
```

## Comparison with RStan

The RStan interface ([**rstan**](https://mc-stan.org/rstan) package) is an
in-memory interface to Stan and relies on R packages like **Rcpp** and
**inline** to call C++ code from R. On the other hand, the CmdStanR interface does
not directly call any C++ code from R, instead relying on CmdStan for
compilation, running algorithms, and writing results to output files.

Both forms of interfacing with Stan have advantages and disadvantages. An
in-memory interface like RStan is able to offer more advanced features than
CmdStanR (for example the `rstan::log_prob()` and `rstan::grad_log_prob()`
methods) but keeping up with Stan releases is more complicated for RStan, often
requiring non-trivial changes to the **rstan** package and requiring new CRAN
releases of **rstan** and **StanHeaders**. On the other hand, with CmdStanR, 
the latest features in Stan will be available from R immediately after updating 
CmdStan, without an update to the **cmdstanr** package. We also anticipate that
running Stan via external processes will have the advantage of playing nicer with 
R (and RStudio) and result in fewer unexpected crashes than when using RStan.

Finally, RStan and CmdStanR have different open source licenses. RStan uses the 
GPL-3 license while the license for CmdStanR (like Stan) is BSD-3, which is a bit 
more permissive.
