---
title: "How does CmdStanR work?"
author: "Jonah Gabry and Rok Češnovar"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{How does CmdStanR work?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r settings-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  # collapse = TRUE,
  dev = "png",
  dpi = 150,
  fig.asp = 0.618,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center",
  comment = NA,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```


## Introduction

This vignette is intended to be read after the [_Getting started with CmdStanR_](http://mc-stan.org/cmdstanr/articles/cmdstanr.html) 
vignette. Please read that first for important background. In this document we
provide additional details about compiling models, passing in data, and how
CmdStan output is saved and read back into R.

We will only use the `$sample()` method in examples, but all model fitting
methods work in a similar way under the hood.

```{r setup, message=FALSE}
library(cmdstanr)
```

## Compilation

### Immediate compilation

The `cmdstan_model()` function creates a new `CmdStanModel` object. The
`CmdStanModel` object stores the path to a Stan program as well as the 
path to a compiled executable: 

```{r start-clean, include=FALSE}
exe <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli")
unlink(exe)
```

```{r compile}
stan_file <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli.stan")
mod <- cmdstan_model(stan_file)
mod$print()
mod$stan_file()
mod$exe_file()
```

Subsequently, if you create a `CmdStanModel` object from the same Stan file
then compilation will be skipped (assuming the file hasn't changed):

```{r already-compiled}
mod <- cmdstan_model(stan_file)
```
 
Internally, `cmdstan_model()` first creates the `CmdStanModel` object from
just the Stan file and then calls its [`$compile()`](http://mc-stan.org/cmdstanr/reference/model-method-compile.html) 
method. Optional arguments to the `$compile()` method can be passed via `...`, 
for example:

```{r compile-options, eval=FALSE}
mod <- cmdstan_model(
  stan_file, 
  force_recompile = TRUE, 
  include_paths = "paths/to/directories/with/included/files", 
  cpp_options = list(stan_threads = TRUE, STANC2 = TRUE)
)
```



### Delayed compilation

It is also possible to delay compilation when creating the `CmdStanModel` object
by specifying `compile=FALSE`. You can later call the `$compile()` method
directly:

```{r compile-method}
unlink(mod$exe_file())
mod <- cmdstan_model(stan_file, compile = FALSE)
mod$exe_file() # not yet created
mod$compile()
mod$exe_file()
```

In all cases the executable is created in the same directory as the file
containing the Stan program.


## Processing data

There are three data formats that CmdStanR allows when fitting a model:

* named list of R objects 
* JSON file
* R dump file

### Named list of R objects

Like the RStan interface, CmdStanR accepts a named list of R objects where the
names correspond to variables declared in the `data` block of the Stan program:

```{r data-list, eval=FALSE}
# data block has 'N' and 'y'
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list)
```

Because CmdStan doesn't accept lists of R objects, CmdStanR will first write the
data to a temporary JSON file using `write_stan_json()`. This happens
internally, but it is also possible to call `write_stan_json()` directly:

```{r write_stan_json}
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)
cat(readLines(json_file), sep = "\n")
```

### JSON file

If you already have your data in a JSON file you can just pass that file
directly to CmdStanR instead of using a list of R objects. For example, we could
pass in the JSON file we created above using `write_stan_json()`:

```{r data-json, eval=FALSE}
fit <- mod$sample(data = json_file)
```


### R dump file

Finally, it is also possible to use the R dump file format. This is *not*
recommended because CmdStan can process JSON faster than R dump, but CmdStanR
allows it because CmdStan will accept files created by `rstan::stan_rdump()`:

```{r data-rdump, eval=FALSE}
rdump_file <- tempfile(fileext = ".data.R")
rstan::stan_rdump(names(data_list), file = rdump_file, envir = list2env(data_list))
cat(readLines(rdump_file), sep = "\n")
fit <- mod$sample(data = rdump_file)
```


## Writing CmdStan output to CSV

### Default temporary files

```{r sample-tempdir, results = "hide"}
data_list <- list(N = 10, y = c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list)
```

When fitting a model, the default behavior is to write the output from CmdStan 
to CSV files in a temporary directory:

```{r output_files}
fit$output_files()
```

These files will be lost if you end your R session or if you remove the 
`fit` object and force (or wait for) garbage collection. 


```{r gc}
files <- fit$output_files()
file.exists(files)

rm(fit)
gc()

file.exists(files)
```

### Non-temporary files

To save these files to a non-temporary location there are two options. You 
can either specify the `output_dir` argument to `mod$sample()` or use 
`fit$save_output_files()` after fitting the model: 


```{r save_output_files, eval=FALSE}
# see ?save_output_files for info on optional arguments
fit$save_output_files(dir = "path/to/directory")
```

```{r output_dir, eval = FALSE}
fit <- mod$sample(
  data = data_list, 
  output_dir = "path/to/directory"
)
```


## Reading CmdStan output into R

### Lazy CSV reading 

With the exception of some diagnostic information, the CSV files are not read
into R until their contents are requested by calling a method that requires them
(e.g., `fit$draws()`, `fit$summary()`, etc.). If we examine the structure of the
`fit` object, notice how the `Private` slot `draws_` is `NULL`, indicating that
the CSV files haven't yet been read into R:

```{r refit, include=FALSE}
fit <- mod$sample(data = data_list)
```
```{r csv-not-read}
str(fit)
```

After we call a method that requires the draws then if we reexamine the 
structure of the object we will see that the `draws_` slot in `Private` 
is no longer empty: 

```{r for-csv-reading}
draws <- fit$draws() # force CSVs to be read into R
str(fit)
```

For models with many parameters, transformed parameters, or generated
quantities, if only some are requested (e.g., by specifying the `variables`
argument to `fit$draws()`) then CmdStanR will only read in the requested
variables (unless they have already been read in).

### read_cmdstan_csv()

Internally, the `read_cmdstan_csv()` function is used to read the CmdStan CSV
files into R. This function is exposed to users, so you can also call it
directly:

```{r read_cmdstan_csv}
# see ?read_cmdstan_csv for info on optional arguments controlling 
# what information is read in
csv_contents <- read_cmdstan_csv(fit$output_files())
str(csv_contents)
```


### Saving and accessing advanced algorithm info (latent dynamics)

If `save_latent_dynamics` is set to `TRUE` when running the `$sample()` method
then additional CSV files are created (one per chain) that provide access to
quantities used under the hood by Stan's implementation of dynamic Hamiltonian
Monte Carlo.

CmdStanR does not yet provide a special method for processing these files but
they can be read into R using R's standard CSV reading functions:


```{r save_latent_dynamics, results = "hide"}
fit <- mod$sample(data = data_list, save_latent_dynamics = TRUE)
```
```{r read-latent-dynamics}
fit$latent_dynamics_files()

# read one of the files in
x <- utils::read.csv(fit$latent_dynamics_files()[1], comment.char = "#")
head(x)
```

The column `lp__` is also provided via `fit$draws()`, and the columns
`accept_stat__`, `stepsize__`, `treedepth__`, `n_leapfrog__`, `divergent__`, and
`energy__` are also provided by `fit$sampler_diagnostics()`, but there are 
several columns unique to the latent dynamics file:

```{r explore-latent-dynamics}
head(x[, c("theta", "p_theta", "g_theta")])
```

Our model has a single parameter `theta` and the three columns above correspond
to `theta` in the _unconstrained_ space (`theta` on the constrained space is
accessed via `fit$draws()`), the auxiliary momentum `p_theta`, and the gradient
`g_theta`. In general, each of these three columns will exist for _every_
parameter in the model.


## Saving fitted model objects

As described above, the contents of the CSV files are only read into R when they
are needed. This means that in order to save a fitted model object containing
_all_ of the posterior draws and sampler diagnostics you should either make sure
to call `fit$draws()` and `fit$sampler_diagnostics()` before saving the object
`fit`, or use the special `$save_object()` method provided by CmdStanR, which
will ensure that everything has been read into R before saving the object using
`saveRDS()`:

```{r save-fit}
temp_rds_file <- tempfile(fileext = ".RDS") # temporary file just for demonstration
fit$save_object(file = temp_rds_file)
```

We can check that this worked by removing `fit` and loading it back in from
the save file:

```{r load-fit}
rm(fit); gc()

fit <- readRDS(temp_rds_file)
fit$summary()
```


