---
title: "Getting started with CmdStanR"
author: "Jonah Gabry and Rok Češnovar"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Getting started with CmdStanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r settings-knitr, include=FALSE}
stopifnot(require(knitr))
opts_chunk$set(
  # collapse = TRUE,
  comment=NA,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE
)
```


## Introduction

CmdStanR is a lightweight interface to [Stan](https://mc-stan.org) for R users
(see [CmdStanPy](https://github.com/stan-dev/cmdstanpy) for Python) that
provides an alternative to the traditional [RStan](https://mc-stan.org/rstan)
interface. See the [*Comparison with RStan*](#comparison-with-rstan) section
later in this vignette for more details on how the two inferfaces differ.

**CmdStanR is unreleased and user facing functions are still subject to change**, 
but the development version can be installed from GitHub:

```{r install_github, eval=FALSE}
devtools::install_github("stan-dev/cmdstanr")
```

CmdStanR (the **cmdstanr** R package) can now be loaded like any other R package:

```{r library, message=FALSE}
library(cmdstanr)
```

## Installing CmdStan

CmdStanR requires a working installation of
[CmdStan](https://mc-stan.org/users/interfaces/cmdstan), the shell interface to
Stan. If you don't have CmdStan installed you can install it yourself or use the
`install_cmdstan()` function provided by CmdStanR.

```{r, include = FALSE}
if (!dir.exists(cmdstan_default_path())) {
  install_cmdstan()
}
```
```{r, eval=FALSE}
install_cmdstan()
```

Before CmdStanR can be used it needs to know where the CmdStan installation is
located. When the package is loaded it tries to help automate this to avoid
having to manually set the path every session:

1. If the environment variable `"CMDSTAN"` exists at load time then its value
will be automatically set as the default path to CmdStan for the R session. This
is useful if your CmdStan installation is not located in the default directory
that would have been used by `install_cmdstan()` (see #2).

2. If no environment variable is found when loaded but the directory
`".cmdstanr/cmdstan"` exists in the user's *home* directory (not working
directory) then it will be set as the path to CmdStan for the R session. This is
the same as the default directory that `install_cmdstan()` uses to install the
latest version of CmdStan, so if that's how you installed CmdStan you shouldn't
need to manually set the path to CmdStan when loading CmdStanR.

If neither of these applies (or you want to subsequently change the path) you
can use the `set_cmdstan_path()` function:

```{r set_cmdstan_path, eval=FALSE}
set_cmdstan_path(PATH_TO_CMDSTAN)
```

If you need to check or get the current path to CmdStan use the 
`cmdstan_path()` function.

## Running CmdStan from R

### Compilation 

The `cmdstan_model()` function creates a new
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html)
object from a file containing a Stan program. Here we'll use the example Stan
program that comes with the CmdStan installation:

```{r cmdstan_model}
file <- file.path(cmdstan_path(), "examples", "bernoulli", "bernoulli.stan")
mod <- cmdstan_model(file)
```

The object `mod` is an [R6](https://r6.r-lib.org/) reference object of class
`CmdStanModel` and behaves similarly to to R's reference class objects and those
in object oriented programming languages. Methods are accessed using the `$`
operator. This design choice allows for CmdStanR and
[CmdStanPy](https://github.com/stan-dev/cmdstanpy) to provide a similar user
experience and share many implementation details, which also helps facilitate
collaboration between developers.

The Stan program can be printed using the `$print()` method: 

```{r compile}
mod$print()  # print the Stan program
```

The path to the compiled executable is available by calling the `$exe_file()`
method:

```{r exe_file}
mod$exe_file()
```

### Model fitting

#### MCMC

To run Stan's default MCMC algorithm we can use the
[`$sample()`](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)
method for
[`CmdStanModel`](https://mc-stan.org/cmdstanr/reference/CmdStanModel.html)
objects. Data can be passed in as a named list of R objects (like for RStan) or
as a path to a data file compatible with CmdStan (R dump or JSON).

```{r sample}
data_list <- list(N = 10, y =c(0,1,0,0,0,0,0,0,0,1))
fit <- mod$sample(data = data_list, seed = 123, num_chains = 2)
```

The
[`$sample()`](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)
method creates [R6](https://r6.r-lib.org/)
[`CmdStanMCMC`](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html)
objects.

```{r summary}
class(fit)
```

The [`$summary()`](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html) and [`$diagnose()`](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html) methods
call CmdStan's `stansummary` and `diagnose` utilities, respectively:

```r
fit$summary()
fit$diagnose()
```

Because CmdStanR is not fully developed, currently the best way to work with the
output is to create a stanfit object (RStan) from the csv output files written
by CmdStan. This can be done by using `rstan::read_stan_csv()` in combination
with the `$output_files()` method of the `CmdStanMCMC` object.

```{r stanfit}
stanfit <- rstan::read_stan_csv(fit$output_files())
print(stanfit)
```

#### Optimization and variational inference

CmdStanR also supports running Stan's optimization algorithms and its algorithms
for variational approximation of full Bayesian inference. These are run via the
`$optimize()` and `$variational()` methods, which are called in a similar way to
the `$sample()` method demonstrated above. For more details follow these links
to the individual documentation pages:

* [`$optimize()`](https://mc-stan.org/cmdstanr/reference/model-method-optimize.html)
* [`$variational()`](https://mc-stan.org/cmdstanr/reference/model-method-variational.html)


## Comparison with RStan

The RStan interface ([**rstan**](https://mc-stan.org/rstan) package) is an
in-memory interface to Stan and relies on R packages like **Rcpp** and
**inline** call C++ code from R. On the other hand, the CmdStanR interface does
not directly call any C++ code from R, instead relying on CmdStan for
compilation, running algorithms, and writing results to output files.

Both forms of interfacing with Stan have advantages and disadvantages. An
in-memory interface like RStan is able to offer more advanced features than
CmdStanR (for example the `rstan::grad_log_prob()` and
`rstan::unconstrain_pars()` methods) but keeping up with Stan releases is more
complicated for RStan, often requiring non-trivial changes to the **rstan**
package and requiring new CRAN releases of **rstan** and **StanHeaders**. With
CmdStanR, the latest features in Stan will be available from R immediately after
updating CmdStan, without an update to the **cmdstanr** package.

Another difference between RStan and CmdStanR is the license. RStan is GPL-3
while the license for CmdStanR (like Stan) is BSD-3, which is a bit more
permissive.

