---
title: "Handling deprecated Stan features with the canonicalizer in CmdStanR"
author: "Rok Češnovar and Jonah Gabry"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Handling deprecated Stan features with the canonicalizer in CmdStanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="children/settings-knitr.Rmd"}
```

## Introduction

This vignette demonstrates how to handle cases where your Stan program contains
deprecated features resulting in deprecation warnings. In most cases, the
Stan-to-C++ compiler can be used to automatically update your code to a
non-deprecated feature that replaces the deprecated one. This vignette showcases
how that automatic conversion can be done using CmdStanR.

The automatic conversion of deprecated features to non-deprecated features is
done using the so-called "canonicalizer", which is part of the Stan-to-C++
compiler. We recommend using CmdStan 2.29.2 or later when using the
canonicalizer and this vignette. The minimum CmdStanR version to run the
code in the vignette is 0.5.0.

```{r library, message=FALSE}
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
```

## Deprecation warnings

The following logistic regression model uses several deprecated language
features, resulting in several warnings during compilation.

```{r logistic}
stan_file <- write_stan_file("
data {
  int<lower=1> k;
  int<lower=0> n;
  matrix[n, k] X;
  int y[n];
}
parameters {
  vector[k] beta;
  real alpha;
}
model {
  # priors
  target += std_normal_log(beta);
  alpha ~ std_normal();

  y ~ bernoulli_logit(X * beta + alpha);
}
")
mod <- cmdstan_model(stan_file)
```


The first warning is about using the deprecated array syntax

```
int y[n];
```

which should be replaced with the new syntax using the `array` keyword:

```
array[n] int y;
```

The second warning is about using the deprecated commenting symbol `#`,
which should be replaced by `//`.

The last warning is about the use of the deprecated `_log` suffix for
probability density and mass functions. In this case the `_log` suffix should be
replaced with `_lpdf`. For probability mass functions the suffix `_lpmf` is
used.

We can go and fix these issues manually or use the canonicalizer as outlined
in the next section.

## Using the canonicalizer

The canonicalizer is available through the `canonicalize` argument of the
`$format()` method of the `CmdStanModel` class. The arguments accepts
`TRUE` and `FALSE` values, in which case all or none of the features of the
canonicalizer are used. It can also accept a list of character vectors that
determine which features of the canonicalizer to use.

The canonincalizer in CmdStan 2.29.2 supports four features: `parentheses`,
`braces`, `includes` and `deprecations`. The `parentheses` and `braces` features
clean up the use of parentheses and braces, while `includes` will replace
`#include` statements with the code from the included files. See the
[canonicalizer section of the Stan User's Guide](https://mc-stan.org/docs/2_29/stan-users-guide/stanc-pretty-printing.html#canonicalizing)
for more details.

In this vignette we will be using the `deprecations` feature that replaces
deprecated Stan model features with non-deprecated ones if possible.

```{r canonicalize, message=FALSE}
mod$format(canonicalize = list("deprecations"))
```

By default, the format function will print the resulting model code. We
can see that all three issues were resolved. `y` is now defined using the
new array keyword, the comment uses `//` and the `std_normal_log()` is replaced
with `std_normal_lpdf()`.

You can also use the `$format()` method to write the updated version of the
model directly to the Stan model file. That can be enabled by setting
`overwrite_file = TRUE`. The previous version of the file will automatically
be backed up to a file with the `.stan.bak` suffix. If that is not desired or
you are using a version system and making a backup is redundant,
you can disable it by setting `backup = FALSE`.

```{r overwrite_file}
mod$format(
    canonicalize = list("deprecations"),
    overwrite_file = TRUE,
    backup = FALSE
)
mod$print()
```
