---
title: "Run Stan on the GPU with OpenCL"
author: "Rok Češnovar and Jonah Gabry"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("CMDSTANR_OPENCL_TESTS"), "true")
vignette: >
  %\VignetteIndexEntry{Using CmdStanR to run Stan on the GPU with OpenCL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette demonstrates how to use the OpenCL capabilites of CmdStan with
CmdStanR. The vignette requires CmdStan 2.26.1 or newer.

## OpenCL runtime

OpenCL is supported on most modern CPUs and GPUs. In order to use
OpenCL in CmdStanR, an OpenCL runtime for the target device must be installed.
A guide for the most common devices is available [here](https://mc-stan.org/docs/2_26/cmdstan-guide/parallelization.html#opencl).

## Compiling a model with OpenCL

By default, models in CmdStanR are compiled without OpenCL support. Once OpenCL
support is enabled, a CmdStan model will make use of OpenCL if the functions
in the model support it. Technically no changes to a model are required to
support OpenCL since the choice of using OpenCL is handled by the compiler,
but it can still be useful to rewrite a model to be more OpenCL-friendly by
using vectorization as much as possible.

Consider a simple logistic regression with parameters `alpha` and `beta`,
covariates `X`, and outcome `y`.

```stan
data {
  int<lower=1> k;
  int<lower=0> n;
  matrix[n, k] X;
  int y[n];
}
parameters {
  vector[k] beta;
  real alpha;
}
model {
  target += std_normal_lpdf(beta);
  target += std_normal_lpdf(alpha);
  target += bernoulli_logit_glm_lpmf(y | X, alpha, beta);
}
```

Some fake data will be useful to run this model:

```{r fake-data, message=FALSE, results='hide'}
library(cmdstanr)

# Generate some fake data
n <- 200000
k <- 20
X <- matrix(rnorm(n * k), ncol = k)

y <- 3 * X[,1] - 2 * X[,2] + 1
p <- runif(n)
y <- ifelse(p < (1 / (1 + exp(-y))), 1, 0)
mdata <- list(k = ncol(X), n = nrow(X), y = y, X = X)
```

In this model, most of the computation will be handled by the
`bernoulli_logit_glm_lpmf` function. Because this is a supported GPU function,
it should be possible to accelerate it with OpenCL. Check 
[here](http://mc-stan.org/math/opencl_support.html) for a list of functions
with OpenCL support.

To build the model with OpenCL support, add
`cpp_options = list(stan_opencl = TRUE)` to the model compile.

```{r compile-opencl, message=FALSE, results='hide'}
# Compile the model with STAN_OPENCL=TRUE
model_cl <- cmdstan_model("opencl-files/bernoulli_logit_glm.stan",
                          cpp_options = list(stan_opencl = TRUE))
```

## Running models with OpenCL

Running models with OpenCL requires specifying the OpenCL platform and device
on which to run the model (there can be multiple). If the system has one GPU
and no OpenCL CPU runtime, the platform and device IDs of the GPU are typically
both `0`, but the `clinfo` tool can be used to figure out for sure what devices
are available.

On an Ubuntu system with both CPU and GPU OpenCL support, `clinfo -l` outputs:

```{bash eval=FALSE}
Platform #0: AMD Accelerated Parallel Processing
 `-- Device #0: gfx906+sram-ecc
Platform #1: Intel(R) CPU Runtime for OpenCL(TM) Applications
 `-- Device #0: Intel(R) Core(TM) i7-4790 CPU @ 3.60GHz
```

The GPU is platform ID 0 and device ID 0, while the CPU is platform ID 1,
device ID 0 . These can be specified with the `opencl_ids` argument 
when running a model. The `opencl_ids` is  supplied as a vector of 
length 2, where the first element is the platform ID and the second
argument is the device ID.

```{r fit-opencl}
fit_cl <- model_cl$sample(data = mdata, chains = 4, parallel_chains = 4,
                          opencl_ids = c(0, 1))
```

Lets run a version without OpenCL:

```{r fit-cpu, message=FALSE}
# no OpenCL version
model <- cmdstan_model("opencl-files/bernoulli_logit_glm.stan")
fit_cpu <- model$sample(data = data_file, chains = 4, parallel_chains = 4,
                    refresh = 0)
```

The speedup of the OpenCL model is:
```{r time-ratio, message=FALSE}
fit_cpu$time()$total/fit_cl$time()$total
```

This speedup will be determined by the GPU/CPU used, the input problem
sizes (data as well as parameters) and if the model uses functions
that can be run on the GPU or other OpenCL device.
