---
title: "Working with Posteriors"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Working with Posteriors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="children/settings-knitr.Rmd"}
```

## Summary

We can easily customise the summary statistics reported by `$summary()` and `$print()`.

```{r}
fit <- cmdstanr::cmdstanr_example("schools", method = "sample")
fit$summary()
```

By default all variables are summaries with the follow functions:
```{r}
posterior::default_summary_measures()
```

To change the variables summarised, we use the variables argument
```{r}
fit$summary(variables = c("mu", "tau"))
```

We can additionally change which functions are used
```{r}
fit$summary(variables = c("mu", "tau"), mean, sd)
```

To summarise all variables with non-default functions, it is necessary to set explicitly set the variables argument, either to `NULL` or the full vector of variable names.
```{r}
fit$metadata()$model_params
fit$summary(variables = NULL, "mean", "median")
```

Summary functions can be specified by character string, function, or using a formula (or anything else supported by [rlang::as_function]). If these arguments are named, those names will be used in the tibble output. If the summary results are named they will take precedence.
```{r}
my_sd <- function(x) c(My_SD = sd(x))
fit$summary(
  c("mu", "tau"), 
  MEAN = mean, 
  "median",
  my_sd,
  ~quantile(.x, probs = c(0.1, 0.9)),
  Minimum = \(x) min(x)
)        
```

Arguments to all summary functions can also be specified with `.args`. 
```{r}
fit$summary(c("mu", "tau"), quantile, .args = list(probs = c(0.025, .05, .95, .975)))
```

The summary functions are applied to the array of sample values, with dimension `iter_sampling`x`chains`.
```{r}
fit$summary(variables = NULL, dim, colMeans)
```

For this reason users may have unexpected results if they use [stats::var()] directly, as it will return a covariance matrix. An alternative is the [distributional::variance] function.
```{r}
fit$summary(c("mu", "tau"), distributional::variance, ~var(as.vector(.x)))
```

Summary functions need not be numeric, but these won't work with `$print()`.

```{r}
strict_pos <- function(x) if (all(x > 0)) "yes" else "no"
fit$summary(variables = NULL, "Strictly Positive" = strict_pos)
# fit$print(variables = NULL, "Strictly Positive" = strict_pos)
```

For more information, see [posterior::summarise_draws()], which is is called by `$summary()`.
