---
title: "Working with Posteriors"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Working with Posteriors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="children/_settings-knitr.Rmd"}
```

```{r include=FALSE}
# Needed temporarily to avoid weird rendering of posterior's tibbles
# in pkgdown sites
print.tbl_df <- function(x, ...) {
  print.data.frame(x)
}
```

## Summary statistics

We can easily customize the summary statistics reported by `$summary()` and `$print()`.

```{r}
fit <- cmdstanr::cmdstanr_example("schools", method = "sample")
fit$summary()
```

By default, all variables are summarized with the following functions:
```{r}
posterior::default_summary_measures()
```

To change the variables summarized, use the `variables` argument:
```{r}
fit$summary(variables = c("mu", "tau"))
```

We can also change which functions are used:
```{r}
fit$summary(variables = c("mu", "tau"), mean, sd)
```

To summarize all variables with non-default functions, it is necessary to
explicitly set the `variables` argument, either to `NULL` or the full vector of
variable names.
```{r}
fit$summary(variables = NULL, "mean", "median")
```

Summary functions can be specified by character string, function, or using a
formula (or anything else supported by `rlang::as_function()`). If these
arguments are named, those names will be used in the tibble output. If the
summary results are named they will take precedence.
```{r}
my_sd <- function(x) c(My_SD = sd(x))
fit$summary(
  c("mu", "tau"), 
  MEAN = mean, 
  "median",
  my_sd,
  ~quantile(.x, probs = c(0.1, 0.9)),
  Minimum = function(x) min(x)
)        
```

Arguments to all summary functions can also be specified with `.args`. 
```{r}
fit$summary(c("mu", "tau"), quantile, .args = list(probs = c(0.025, .05, .95, .975)))
```

Each summary function is applied separately to each variable and receives a
matrix whose rows are saved iterations and whose columns are chains.
```{r}
fit$summary(variables = NULL, dim, colMeans)
```


For this reason users may have unexpected results if they use `stats::var()` directly, as it will return a covariance matrix. An alternative is the `distributional::variance()` function, 
which can also be accessed via `posterior::variance()`.
```{r}
fit$summary(c("mu", "tau"), posterior::variance, ~var(as.vector(.x)))
```

Summary functions need not return numeric values when used with `$summary()`.
The `$print()` method requires numeric summary columns because it rounds them
to the requested number of digits.

```{r}
strict_pos <- function(x) if (all(x > 0)) "yes" else "no"
fit$summary(variables = NULL, "Strictly Positive" = strict_pos)
# fit$print(variables = NULL, "Strictly Positive" = strict_pos)
```

For more information, see `posterior::summarise_draws()`, which is called by `$summary()`.


## Extracting posterior draws/samples

The [`$draws()`](https://mc-stan.org/cmdstanr/reference/fit-method-draws.html)
method extracts draws in formats provided by the
[**posterior**](https://mc-stan.org/posterior/) package. The
[_Getting started with CmdStanR_](https://mc-stan.org/cmdstanr/articles/cmdstanr.html#extracting-draws)
vignette introduces the most commonly used formats and how to convert between
them.

```{r draws, message=FALSE}
fit$draws("mu")
fit$draws("theta")
fit$draws(c("mu", "theta[1]"), format = "df")
```

For MCMC fits, `inc_warmup = TRUE` includes warmup draws, but only if
`save_warmup = TRUE` was specified when fitting the model.

For more ways to manipulate draws, see the **posterior** package
[vignettes](https://mc-stan.org/posterior/articles/index.html) and
[documentation](https://mc-stan.org/posterior/reference/index.html).

### Structured draws similar to `rstan::extract()`

The **posterior** package provides two useful ways to work with variables while
preserving their original dimensions.

`posterior::extract_list_of_variable_arrays()` returns a named list containing
one array per variable. Setting `with_chains = FALSE` combines the chains,
giving the same general structure as the list returned by `rstan::extract()`:

```{r variable-arrays}
draw_arrays <- posterior::extract_list_of_variable_arrays(
  fit$draws(),
  variables = c("mu", "theta"),
  with_chains = FALSE
)
str(draw_arrays)
```

The first dimension of each array indexes draws, and any remaining dimensions
match the dimensions of the corresponding Stan variable.

Alternatively, the **posterior** package's `rvar` format represents each
variable as a multidimensional random variable, with its posterior draws
handled behind the scenes:

```{r structured-draws}
draws_rvars <- posterior::as_draws_rvars(
  fit$draws(c("mu", "theta"))
)
theta_rvar <- draws_rvars$theta

# Compute the difference for every draw using natural vector indexing
# The posterior draws are handled automatically
theta_difference <- theta_rvar[1] - theta_rvar[2]
theta_difference

hist(
  posterior::draws_of(theta_difference),
  main = "Difference between theta[1] and theta[2]",
  xlab = "theta[1] - theta[2]"
)

# Direct access to the underlying draws is also available
theta_array <- posterior::draws_of(theta_rvar)
dim(theta_array)
```

The object `theta_rvar` behaves like the vector declared in the Stan program.
`theta_array` provides direct access to its underlying draws, with the first
dimension indexing draws. See the
[`rvar` vignette](https://mc-stan.org/posterior/articles/rvar.html) for details.
