---
title: "Working with Posteriors"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Working with Posteriors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="children/_settings-knitr.Rmd"}
```

<!-- 
NOTE: the hack below of using print.data.frame in chunks with echo=FALSE 
is used because the pillar formatting of posterior draws_summary objects
isn't playing nicely with pkgdown::build_articles(). When that is fixed
using options(digits=2) won't be necessary anymore.
-->
```{r, include=FALSE}
options(digits=2)
```

## Summary

We can easily customise the summary statistics reported by `$summary()` and `$print()`.

```{r eval=FALSE}
fit <- cmdstanr::cmdstanr_example("schools", method = "sample")
fit$summary()
```
```{r echo=FALSE}
fit <- cmdstanr::cmdstanr_example("schools", method = "sample")
print.data.frame(fit$summary())
```

By default all variables are summaries with the follow functions:
```{r}
posterior::default_summary_measures()
```

To change the variables summarised, we use the variables argument
```{r eval=FALSE}
fit$summary(variables = c("mu", "tau"))
```
```{r echo=FALSE}
print.data.frame(fit$summary(variables = c("mu", "tau")))
```

We can additionally change which functions are used
```{r eval=FALSE}
fit$summary(variables = c("mu", "tau"), mean, sd)
```
```{r echo=FALSE}
print.data.frame(fit$summary(variables = c("mu", "tau"), mean, sd))
```

To summarise all variables with non-default functions, it is necessary to set explicitly set the variables argument, either to `NULL` or the full vector of variable names.
```{r eval=FALSE}
fit$metadata()$model_params
fit$summary(variables = NULL, "mean", "median")
```
```{r echo=FALSE}
fit$metadata()$model_params
print.data.frame(fit$summary(variables = NULL, "mean", "median"))
```

Summary functions can be specified by character string, function, or using a formula (or anything else supported by [rlang::as_function]). If these arguments are named, those names will be used in the tibble output. If the summary results are named they will take precedence.
```{r eval=FALSE}
my_sd <- function(x) c(My_SD = sd(x))
fit$summary(
  c("mu", "tau"), 
  MEAN = mean, 
  "median",
  my_sd,
  ~quantile(.x, probs = c(0.1, 0.9)),
  Minimum = function(x) min(x)
)        
```
```{r echo=FALSE}
my_sd <- function(x) c(My_SD = sd(x))
print.data.frame(fit$summary(
  c("mu", "tau"), 
  MEAN = mean, 
  "median",
  my_sd,
  ~quantile(.x, probs = c(0.1, 0.9)),
  Minimum = function(x) min(x)
))
```


Arguments to all summary functions can also be specified with `.args`. 
```{r eval=FALSE}
fit$summary(c("mu", "tau"), quantile, .args = list(probs = c(0.025, .05, .95, .975)))
```
```{r echo=FALSE}
print.data.frame(fit$summary(c("mu", "tau"), quantile, .args = list(probs = c(0.025, .05, .95, .975))))
```

The summary functions are applied to the array of sample values, with dimension `iter_sampling`x`chains`.
```{r eval=FALSE}
fit$summary(variables = NULL, dim, colMeans)
```
```{r echo=FALSE}
print.data.frame(fit$summary(variables = NULL, dim, colMeans))
```


For this reason users may have unexpected results if they use `stats::var()` directly, as it will return a covariance matrix. An alternative is the `distributional::variance()` function, 
which can also be accessed via `posterior::variance()`.
```{r eval=FALSE}
fit$summary(c("mu", "tau"), posterior::variance, ~var(as.vector(.x)))
```
```{r echo=FALSE}
print.data.frame(fit$summary(c("mu", "tau"), posterior::variance, ~var(as.vector(.x))))
```


Summary functions need not be numeric, but these won't work with `$print()`.

```{r eval=FALSE}
strict_pos <- function(x) if (all(x > 0)) "yes" else "no"
fit$summary(variables = NULL, "Strictly Positive" = strict_pos)
# fit$print(variables = NULL, "Strictly Positive" = strict_pos)
```
```{r echo=FALSE}
strict_pos <- function(x) if (all(x > 0)) "yes" else "no"
print.data.frame(fit$summary(variables = NULL, "Strictly Positive" = strict_pos))
# fit$print(variables = NULL, "Strictly Positive" = strict_pos)
```

For more information, see `posterior::summarise_draws()`, which is called by `$summary()`.
