---
title: "Running Stan on the GPU with OpenCL"
author: "Rok Češnovar and Jonah Gabry"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true") && isTRUE(Sys.getenv("CMDSTANR_OPENCL_TESTS") %in% c("true", "1"))
vignette: >
  %\VignetteIndexEntry{Running Stan on the GPU with OpenCL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r child="../children/_settings-knitr.Rmd"}
```

## Introduction

This vignette demonstrates how to use the OpenCL capabilities of CmdStan with
CmdStanR. The functionality described in this vignette requires CmdStan 2.26.1
or newer.

As of version 2.26.1, users can expect speedups with OpenCL when using vectorized
probability distribution functions (functions with the `_lpdf` or `_lpmf`
suffix) and when the input variables contain at least 20,000 elements.

The actual speedup for a model will depend on the particular `lpdf/lpmf`
functions used and whether the `lpdf/lpmf` functions are the bottlenecks of the
model. The more computationally complex the function is, the larger the expected
speedup. The biggest speedups are expected when using the specialized GLM
functions.

In order to establish the bottlenecks in your model we recommend using
[profiling](https://mc-stan.org/cmdstanr/articles/profiling.html),
which was introduced in Stan version 2.26.0.

## OpenCL runtime

OpenCL is supported on most modern CPUs and GPUs. In order to use
OpenCL in CmdStanR, an OpenCL runtime for the target device must be installed.
A guide for the most common devices is available in the CmdStan manual's
[chapter on parallelization](https://mc-stan.org/docs/2_26/cmdstan-guide/parallelization.html#opencl).

In case of using Windows, CmdStan requires the `OpenCL.lib` to compile the model.
If you experience issue compiling the model with OpenCL, run the below script and set
`path_to_opencl_lib` to the path to the `OpenCL.lib` file on your system.
If you are using CUDA, the path should be similar to the one listed here.

```{r windows-opencl, eval=FALSE}
path_to_opencl_lib <- "C:/Program Files/NVIDIA GPU Computing Toolkit/CUDA/v11.5/lib/x64"
cpp_options = list(
  paste0("LDFLAGS+= -L\"",path_to_opencl_lib,"\" -lOpenCL")
)

cmdstanr::cmdstan_make_local(cpp_options = cpp_options)
cmdstanr::rebuild_cmdstan()
```

## Compiling a model with OpenCL

By default, models in CmdStanR are compiled *without* OpenCL support. Once OpenCL
support is enabled, a CmdStan model will make use of OpenCL if the functions
in the model support it. Technically no changes to a model are required to
support OpenCL since the choice of using OpenCL is handled by the compiler,
but it can still be useful to rewrite a model to be more OpenCL-friendly by
using vectorization as much as possible when using probability distributions.

Consider a simple logistic regression with parameters `alpha` and `beta`,
covariates `X`, and outcome `y`.

```
data {
  int<lower=1> k;
  int<lower=0> n;
  matrix[n, k] X;
  int y[n];
}
parameters {
  vector[k] beta;
  real alpha;
}
model {
  target += std_normal_lpdf(beta);
  target += std_normal_lpdf(alpha);
  target += bernoulli_logit_glm_lpmf(y | X, alpha, beta);
}
```

Some fake data will be useful to run this model:

```{r fake-data, message=FALSE, results='hide'}
library(cmdstanr)

# Generate some fake data
n <- 250000
k <- 20
X <- matrix(rnorm(n * k), ncol = k)
y <- rbinom(n, size = 1, prob = plogis(3 * X[,1] - 2 * X[,2] + 1))
mdata <- list(k = k, n = n, y = y, X = X)
```

In this model, most of the computation will be handled by the
`bernoulli_logit_glm_lpmf` function. Because this is a supported GPU function,
it should be possible to accelerate it with OpenCL. Check
[here](http://mc-stan.org/math/d0/d8f/opencl_support.html) for a list of functions
with OpenCL support.

To build the model with OpenCL support, add
`cpp_options = list(stan_opencl = TRUE)` at the compilation step.

```{r compile-opencl, message=FALSE, results='hide'}
# Compile the model with STAN_OPENCL=TRUE
mod_cl <- cmdstan_model("opencl-files/bernoulli_logit_glm.stan",
                        cpp_options = list(stan_opencl = TRUE))
```

## Running models with OpenCL

Running models with OpenCL requires specifying the OpenCL platform and device
on which to run the model (there can be multiple). If the system has one GPU
and no OpenCL CPU runtime, the platform and device IDs of the GPU are typically
both `0`, but the `clinfo` tool can be used to figure out for sure which devices
are available.

On an Ubuntu system with both CPU and GPU OpenCL support, `clinfo -l` outputs:

```
Platform #0: AMD Accelerated Parallel Processing
 `-- Device #0: gfx906+sram-ecc
Platform #1: Intel(R) CPU Runtime for OpenCL(TM) Applications
 `-- Device #0: Intel(R) Core(TM) i7-4790 CPU @ 3.60GHz
```

On this system the GPU is platform ID 0 and device ID 0, while the CPU is
platform ID 1, device ID 0. These can be specified with the `opencl_ids`
argument when running a model. The `opencl_ids` is  supplied as a vector of
length 2, where the first element is the platform ID and the second argument is
the device ID.

```{r fit-opencl}
fit_cl <- mod_cl$sample(data = mdata, chains = 4, parallel_chains = 4,
                        opencl_ids = c(0, 0), refresh = 0)
```

We'll also run a version without OpenCL and compare the run times.

```{r fit-cpu, message=FALSE}
# no OpenCL version
mod <- cmdstan_model("opencl-files/bernoulli_logit_glm.stan", force_recompile = TRUE)
fit_cpu <- mod$sample(data = mdata, chains = 4, parallel_chains = 4, refresh = 0)
```

The speedup of the OpenCL model is:
```{r time-ratio, message=FALSE}
fit_cpu$time()$total / fit_cl$time()$total
```

This speedup will be determined by the particular GPU/CPU used, the input
problem sizes (data as well as parameters) and if the model uses functions that
can be run on the GPU or other OpenCL devices.
